---
title: "Deep-Sea Glass Sponge Community Ecology in the Hawaiian Archipelago"
author: "Diego Johansen, Marine Scientist, Kailua-Kona, Hawai'i"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: kable
    latex_engine: xelatex
    highlight: tango
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    self_contained: true
    code_folding: show
    df_print: paged
    fig_caption: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center",
  fig.width = 7.5, fig.height = 5,
  out.width = "100%",   # scale every figure to the page/column width (prevents PDF clipping)
  dpi = 150
)

# Keep wide tables from overflowing the PDF page margins.
options(knitr.kable.NA = "")
if (knitr::is_latex_output()) {
  knitr::opts_chunk$set(fig.width = 6.5, fig.height = 4.2)
}
```

```{r libraries}
library(tidyverse)
library(stringr)
library(forcats)
library(vegan)
library(GGally)
library(viridis)
library(patchwork)
library(scales)
```

# Summary {-}

Glass sponges (class Hexactinellida) are deep-sea filter feeders that build intricate skeletons of
silica. Those skeletons form biogenic structure on the seafloor that other deep-sea organisms rely
on for shelter and settlement. Because the animals occur at depths that are rarely surveyed, most of
what is known about them comes from opportunistic remotely operated vehicle (ROV) and submersible
dives rather than from systematic sampling programs.

This report analyzes 964 such observations from around the Main Hawaiian Islands, collected by
NOAA's Office of Ocean Exploration and Research. The central question is simple to state but
difficult to answer rigorously: which environmental and geographic factors determine where
particular glass sponges occur?

The most strongly supported answer is depth. Individual species occupy distinct depth ranges, and
that pattern persists across every analysis applied here. A secondary question — whether sponge
communities differ from one island to another — is treated more cautiously, because island identity
is confounded with how each dive was conducted: the equipment used, the amount of search effort, and
the specific locations visited. Without accounting for these factors, ordinary differences in
sampling could be mistaken for genuine ecological ones.

The analysis is presented in two stages. The first describes the structure of the dataset directly.
The second is more rigorous: it treats each dive as the unit of analysis, corrects for the large
differences in how much each dive recorded, and subjects every conclusion to a set of checks before
accepting it. The aim throughout is to be explicit about what these data can and cannot demonstrate.

A practical note: the R code is shown for transparency, but it can be collapsed with the *Code*
buttons at the right of each block for easier reading.

# The Dataset

Each record in this dataset represents a single glass sponge observed during a deep-sea dive in the
Hawaiian Islands. NOAA compiled the records across several expeditions conducted with ROVs and
crewed submersibles, yielding 964 observations in total. For each observation, the dataset documents
the organism's identity, the depth at which it was seen, and its location.

Specifically, each record includes:

- the scientific name and the taxonomic rank to which the sponge was identified (ranging from genus to subspecies)
- the depth of observation, in meters
- geographic coordinates (latitude and longitude)
- a locality descriptor (an island, a seamount, or a named feature)
- dive identifiers (Station, EventID, and SurveyID, which records the expedition), along with the sampling equipment used

The dive identifiers are more consequential than they first appear. Because they link each
observation to a specific dive, they make it possible to treat the dive — rather than the individual
sponge — as the unit of analysis. This is a key methodological point: aggregating to the dive level
is what keeps the statistical tests later in the report valid rather than misleading.

```{r load-data}
data <- read_csv("data/glass_sponges.csv")
```

# Data Cleaning and Preparation

The raw data require several preparation steps before analysis. Locality names were recorded
inconsistently across expeditions, so the first task is to standardize them into consistent
island-level labels. The second is to retain the dive-level metadata (Station, sampling gear, and
expedition), which the inferential analysis depends on heavily. Finally, observations identified only
to coarse taxonomic ranks (class or order) are removed, so that the community comparisons are made at
a consistent level of taxonomic resolution. The specific steps are as follows:

1. Create a lookup table for irregular locality names that don't follow a standard pattern.
2. Remove administrative/metadata columns not needed for ecological analysis.
3. Filter out off-scope seamount sites (Ellis, McCall) outside the Hawaiian Islands focus area.
4. Standardize locality names to island-level identifiers.
5. Retain only observations identified to genus level or finer (genus, subgenus, family, subfamily, species, subspecies), excluding records identified only to class or order.

```{r clean-data}
specific_to_island <- c(
  "Mahukona fissure"          = "Mahukona fissure, Hawaii",
  "Mahukona 900m Terrace"     = "Mahukona 900m Terrace, Hawaii",
  "Mahukona 900m Terrace II"  = "Mahukona 900m Terrace II, Hawaii",
  "Canyon Head 2 Kohala"      = "Canyon Head 2 Kohala, Hawaii",
  "Kohala Plunge Pool I"      = "Kohala Plunge Pool I, Hawaii",
  "Waianae Landslide"         = "Waianae Landslide, Oahu",
  "Kahoolawe Rift I"          = "Kahoolawe",
  "Kahoolawe Rift II"         = "Kahoolawe",
  "Plunge Pool Molokai"       = "Molokai",
  "Kauai Cones II"            = "Kauai"
)

cleaned_sponges <- data %>%
  # Keep dive-level identifiers (Station, EventID, SurveyID), sampling gear, and
  # date so we can use the dive as the sampling unit and account for effort.
  select(-ShallowFlag, -DatasetID, -CatalogNumber, -SampleID, -ImageURL,
         -Repository, -VernacularNameCategory, -IdentificationQualifier,
         -DepthMethod, -RecordType, -DataProvider, -`LocationAccuracy (m)`) %>%
  filter(!Locality %in% c("Hawaiian Archipelago, Ellis Seamount",
                           "Hawaiian Archipelago, McCall Seamount")) %>%
  mutate(Locality = case_when(
    str_detect(Locality, "^Main Hawaiian Islands(, Hawaii Island,| \\| Hawaii \\|)\\s*") ~
      str_replace(Locality, "^Main Hawaiian Islands(, Hawaii Island,| \\| Hawaii \\|)\\s*", "") %>%
      paste0(", Hawaii"),
    str_detect(Locality, "^Main Hawaiian Islands(, Oahu Island,| \\| Oahu \\|)\\s*") ~
      str_replace(Locality, "^Main Hawaiian Islands(, Oahu Island,| \\| Oahu \\|)\\s*", "") %>%
      paste0(", Oahu"),
    Locality %in% names(specific_to_island) ~ specific_to_island[Locality],
    TRUE ~ Locality
  )) %>%
  # Derive a clean island-level grouping from the standardized locality string.
  mutate(Island = case_when(
    str_detect(Locality, regex("oahu", ignore_case = TRUE))      ~ "Oahu",
    str_detect(Locality, regex("kahoolawe", ignore_case = TRUE)) ~ "Kahoolawe",
    str_detect(Locality, regex("molokai", ignore_case = TRUE))   ~ "Molokai",
    str_detect(Locality, regex("kauai", ignore_case = TRUE))     ~ "Kauai",
    str_detect(Locality, regex("maui", ignore_case = TRUE))      ~ "Maui",
    str_detect(Locality, regex("hawaii|mahukona|kohala|south pt", ignore_case = TRUE)) ~ "Hawaii",
    TRUE ~ "Other"
  )) %>%
  filter(TaxonRank %in% c("genus", "subgenus", "family", "subfamily", "subspecies", "species"))

summary(cleaned_sponges)
```

The summary above serves as an integrity check on the cleaned dataset. It confirms that depths and
coordinates fall within plausible ranges, that key fields contain no unexpected missing values, and
that the island grouping was assigned correctly. With the data validated, the following sections
describe its structure in detail.

# Summary Statistics

Before modeling, it is useful to characterize the cleaned dataset directly. The tables below
summarize the observations along several dimensions — taxonomic resolution, species frequency,
locality, and depth. Two patterns emerge early and recur throughout the analysis: glass sponges are
strongly structured by depth, and the dives that produced these records varied substantially in
their sampling effort.

**Table 1. Records by taxonomic rank.** The level of identification varies considerably, from
subspecies down to genus. This inconsistency is worth noting at the outset, because it could
influence community comparisons; later, a sensitivity analysis collapses all identifications to genus
to test whether the conclusions depend on taxonomic resolution.

```{r tax-rank}
cleaned_sponges %>%
  count(TaxonRank, name = "n_observations") %>%
  arrange(desc(n_observations))
```

**Table 2. The ten most frequently observed taxa.** A small number of taxa account for the majority
of records, while many others appear only a handful of times. This long tail of rarely observed
species is one reason for caution in the later statistical tests: there is limited information
available for the uncommon taxa.

```{r top-species}
cleaned_sponges %>%
  count(ScientificName, sort = TRUE, name = "n_observations") %>%
  head(10)
```

**Table 3. Observations per locality.** The records are distributed unevenly across sites, with a
few localities contributing most of the observations. This imbalance in sampling effort is the
primary reason the analysis relies on the dive as the unit of analysis rather than pooling all
observations together.

```{r obs-per-locality}
cleaned_sponges %>%
  count(Locality, sort = TRUE, name = "n_observations")
```

**Table 4. Depth ranges of the most-observed taxa.** Mean, minimum, and maximum depth for the most
frequently recorded taxa. Taxa whose depth ranges show little overlap provide the first indication
of depth partitioning — different species occupying distinct portions of the water column.

```{r depth-by-species}
cleaned_sponges %>%
  group_by(ScientificName) %>%
  summarise(
    mean_depth = round(mean(`DepthInMeters (m)`, na.rm = TRUE), 1),
    min_depth  = min(`DepthInMeters (m)`, na.rm = TRUE),
    max_depth  = max(`DepthInMeters (m)`, na.rm = TRUE),
    n_obs      = n(),
    .groups    = "drop"
  ) %>%
  arrange(desc(n_obs)) %>%
  slice_head(n = 10)
```

**Table 5. Depth range and sampling effort by locality.** This combines where sampling occurred with
how deep and how intensively each site was surveyed. It is useful for identifying sites that were
both shallow and lightly sampled, which warrant particular caution in interpretation.

```{r depth-by-locality}
cleaned_sponges %>%
  group_by(Locality) %>%
  summarise(
    mean_depth = round(mean(`DepthInMeters (m)`, na.rm = TRUE), 1),
    min_depth  = min(`DepthInMeters (m)`, na.rm = TRUE),
    max_depth  = max(`DepthInMeters (m)`, na.rm = TRUE),
    n_obs      = n(),
    .groups    = "drop"
  ) %>%
  arrange(desc(n_obs))
```

**Table 6. Species richness by locality.** The number of distinct taxa recorded at each site. This
raw richness is deliberately not yet interpreted ecologically: sites with more observations tend to
show higher richness simply because they were sampled more intensively. The inferential section
corrects for this using rarefaction.

```{r richness-by-locality}
cleaned_sponges %>%
  group_by(Locality) %>%
  summarise(species_richness = n_distinct(ScientificName), .groups = "drop") %>%
  arrange(desc(species_richness))
```

# Exploratory Visualizations

The following four figures visualize the patterns suggested by the summary tables. They are
descriptive rather than confirmatory: they show where and at what depths sponges were observed, but
they do not yet account for differences in sampling effort between dives. They are best read as
orientation for the formal analysis that follows.

## Mean depth by species

This figure provides a first look at depth partitioning. If species occupy genuinely different
depths, their mean values should spread across a wide range rather than cluster around a single
value.

```{r chart1-mean-depth, fig.cap="Mean observed depth for each glass sponge taxon, ordered from deepest to shallowest. The spread of values across taxa is the first visual evidence that different species occupy different depth bands."}
species_depth_summary <- cleaned_sponges %>%
  group_by(ScientificName) %>%
  summarise(mean_depth = mean(`DepthInMeters (m)`, na.rm = TRUE)) %>%
  arrange(desc(mean_depth))

ggplot(species_depth_summary,
       aes(x = reorder(ScientificName, mean_depth), y = mean_depth, fill = ScientificName)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(round(mean_depth, 0), " m")),
            hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_y_reverse(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "Mean Depth of Glass Sponges by Species",
    x = "Scientific Name",
    y = "Mean Depth (m)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    plot.title.position = "plot",
    axis.text.y = element_text(face = "italic")
  )
```

## Overall depth distribution

This histogram shows how observations are distributed across the water column. Its shape indicates
which depths the dataset can characterize with confidence and which are only sparsely represented.

```{r chart2-depth-histogram, fig.cap="Distribution of all glass sponge observations across depth. Most records concentrate in the upper part of the surveyed range, with progressively fewer observations at greater depths."}
ggplot(cleaned_sponges, aes(x = `DepthInMeters (m)`)) +
  geom_histogram(bins = 30, fill = "#1f78b4", color = "white", alpha = 0.9) +
  labs(
    title = "Depth Distribution of Glass Sponge Observations",
    x = "Depth (m)",
    y = "Number of Observations"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold")
  )
```

## Depth distribution by site

Each point represents one observation. Beyond differences in the number of sponges recorded, sites
also differ in the depth range each one covered. This is an early indication that locality and depth
effects will be difficult to separate fully — a theme the inferential analysis addresses directly.

```{r chart3-depth-by-site, fig.cap="Depth of every observation, grouped by site. Differences in both the number of points and their depth spread reflect how unevenly the sites were sampled."}
ggplot(cleaned_sponges, aes(x = Locality, y = `DepthInMeters (m)`, color = Locality)) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +
  labs(
    title = "Depth Distribution of Glass Sponges by Site",
    x = "Site",
    y = "Depth (m)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y = element_text(size = 10),
    legend.position = "none"
  )
```

## Depth ranges by species

This figure offers the clearest exploratory evidence of depth partitioning. Where the boxes sit at
distinct heights with limited overlap, those species are occupying separate depth bands rather than
ranging freely throughout the water column.

```{r chart4-depth-boxplot, fig.cap="Depth range (box) and individual observations (points) for each taxon. Non-overlapping boxes indicate species partitioning the water column by depth — the central pattern the inferential analysis goes on to test formally."}
ggplot(cleaned_sponges,
       aes(x = fct_reorder(ScientificName, `DepthInMeters (m)`), y = `DepthInMeters (m)`)) +
  geom_boxplot(fill = "steelblue", color = "black", alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1.8, color = "darkblue") +
  labs(
    title = "Depth Ranges of Glass Sponges by Species",
    x = "Scientific Name",
    y = "Depth (m)"
  ) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(face = "italic", size = 10),
    axis.title = element_text(face = "bold")
  )
```

## Depth distribution by species (ridgelines)

The boxplot above summarizes each species' depth range; this figure shows the full *shape* of each
distribution. Every species is drawn as a smoothed density curve along the depth axis and stacked in
order of median depth, so depth partitioning appears as a staircase of curves shifting steadily
deeper. Only taxa with at least five observations are shown, since density estimates are unreliable
for very few points. (This figure requires the `ggridges` package; it is skipped automatically if the
package is not installed.)

```{r ridgeline, fig.height=6.5, fig.cap="Smoothed depth distributions for each taxon with at least five observations, ordered by median depth and shaded by depth. The progressive downward shift of the curves illustrates depth partitioning among species."}
if (requireNamespace("ggridges", quietly = TRUE)) {
  ridge_dat <- cleaned_sponges %>%
    add_count(ScientificName, name = "n_sp") %>%
    filter(n_sp >= 5)

  ggplot(ridge_dat,
         aes(x = `DepthInMeters (m)`,
             y = fct_reorder(ScientificName, `DepthInMeters (m)`, .fun = median),
             fill = after_stat(x))) +
    ggridges::geom_density_ridges_gradient(scale = 2.2, rel_min_height = 0.01,
                                           colour = "white", linewidth = 0.3) +
    scale_fill_viridis(option = "mako", direction = -1, name = "Depth (m)") +
    labs(title = "Depth Distribution of Glass Sponges by Species",
         x = "Depth (m)", y = NULL) +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold"),
          axis.text.y = element_text(face = "italic"))
} else {
  message("Install the 'ggridges' package to render the depth ridgeline plot.")
}
```

# Ordination: Visualizing Community Structure

The next analyses address a more complex question: which localities support similar assemblages of
sponges, and which differ? The standard tool for this is ordination. The principle is more intuitive
than the terminology suggests: a large table recording which species occurred where, and in what
numbers, is reduced to a two-dimensional map on which localities with similar communities appear
close together and dissimilar localities appear far apart. Two complementary methods are used here —
Non-metric Multidimensional Scaling (NMDS) and Principal Coordinates Analysis (PCoA) — which are two
established ways of producing such a map.

Both methods begin from a single measure of how *different* any two localities' communities are: the
**Bray–Curtis dissimilarity**, the conventional choice in community ecology. For a pair of sites it
ranges from 0 to 1. A value of 0 means the two sites have identical communities — the same taxa in
the same proportions — while a value of 1 means they share no taxa at all. It is computed by
comparing the abundance of every taxon between the two sites, summing the differences, and scaling by
the combined total, so that both *which* taxa are present and *how abundant* each one is contribute to
the result. Bray–Curtis is well suited to ecological data because it ignores joint absences (two
sites are not considered similar merely because many taxa are missing from both) and because it
responds to differences in abundance rather than presence alone. Every ordination and statistical
test in this section is built on this measure.

> An important caveat applies to this section. This first ordination groups the data by *locality*
> and uses raw observation counts, so a more heavily sampled site appears more abundant even when it
> is not. With only a handful of localities, statistical power is also limited. These plots should
> therefore be read as an exploratory sketch. The rigorous version — one point per dive, with effort
> accounted for and claims formally tested — follows in the next section.

## Building the community matrix

```{r build-community-matrix}
top_localities <- cleaned_sponges %>%
  count(Locality) %>%
  slice_max(n, n = 10) %>%
  pull(Locality)

filtered_sponges <- cleaned_sponges %>%
  filter(Locality %in% top_localities) %>%
  filter(Locality != "Mahukona fissure, Hawaii")

community_matrix <- filtered_sponges %>%
  count(Locality, ScientificName) %>%
  pivot_wider(names_from = ScientificName, values_from = n, values_fill = 0) %>%
  column_to_rownames("Locality") %>%
  .[rowSums(.) > 0, ]

dim(community_matrix)
```

## NMDS ordination

```{r nmds}
nmds <- metaMDS(community_matrix, distance = "bray", k = 2, trymax = 100)
print(nmds)
stressplot(nmds, main = "Shepard Plot for NMDS (Bray-Curtis)")
```

## NMDS goodness of fit

The "stress" value reported by the ordination quantifies how much information is lost when the
community data are compressed into two dimensions. Lower values indicate a better representation; by
convention, stress below approximately 0.2 is considered an acceptable fit.

```{r nmds-goodness}
site_goodness <- goodness(nmds)
site_scores_base <- scores(nmds, display = "sites")

plot(nmds, type = "n", main = "NMDS Goodness of Fit")
points(site_scores_base,
       cex = rescale(site_goodness, to = c(1, 6)),
       pch = 21, bg = "skyblue")
text(site_scores_base, labels = rownames(site_scores_base), cex = 0.7, pos = 3)
```

## PCoA and scree plot

Principal Coordinates Analysis (PCoA) provides an alternative ordination of the same dissimilarity
matrix. The accompanying scree plot shows the proportion of total community variation captured by
each ordination axis, indicating whether the first few axes account for most of the structure.

```{r pcoa-scree}
bray_dist   <- vegdist(community_matrix, method = "bray")
pcoa        <- cmdscale(bray_dist, eig = TRUE, k = nrow(community_matrix) - 1)
eig_vals    <- pcoa$eig
var_explained <- eig_vals / sum(eig_vals[eig_vals > 0]) * 100

scree_df <- data.frame(
  Axis     = paste0("PCo", seq_along(var_explained)),
  Variance = var_explained
)

ggplot(scree_df[1:10, ], aes(x = Axis, y = Variance)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(Variance, 1), "%")), vjust = -0.5, size = 4) +
  labs(
    title = "Scree Plot of PCoA (Bray-Curtis)",
    x = "Principal Coordinate Axis",
    y = "Variance Explained (%)"
  ) +
  theme_minimal(base_size = 11)
```

## NMDS with depth and island overlaid

This figure presents the same NMDS ordination, annotated so the ecological patterns are easier to
read. Several elements guide interpretation:

- **Position and distance.** Each point is a locality, and the spacing between points reflects
  Bray–Curtis dissimilarity: points close together support similar sponge communities, while points
  far apart share few taxa or very different proportions. Only relative distances carry meaning — the
  axes have no units, and the entire configuration can be rotated or flipped without changing its
  interpretation.
- **Color.** Points are shaded by each locality's mean sampling depth. If depth structures the
  communities, the colors should grade smoothly across the ordination rather than being scattered at
  random — a visual counterpart to the formal depth tests later in the report.
- **Island hulls.** Localities belonging to the same island are enclosed in a shaded polygon. Hulls
  that occupy separate regions with little overlap would indicate that islands support distinct
  communities; substantially overlapping hulls indicate they do not. Given the small number of
  localities, this should be read as suggestive rather than conclusive.
- **Species vectors.** The green arrows point toward the taxa most strongly associated with each
  region of the map; a locality positioned in the direction of a given arrow tends to be relatively
  rich in that taxon.

As with the other plots in this section, this is the locality-level, effort-naive version. The
formally tested, dive-level ordination — with environmental vectors fitted and significance assessed
— appears later in the inferential analysis.

```{r nmds-biplot, fig.cap="Exploratory NMDS ordination of localities (Bray-Curtis). Points are sites colored by mean depth; arrows are species; hulls group localities by island. Read distances between points as community similarity. Exploratory only — see the inferential section for tested versions."}
site_scores <- scores(nmds, display = "sites") %>%
  as.data.frame() %>%
  rownames_to_column("Locality")

site_meta <- cleaned_sponges %>%
  filter(Locality %in% site_scores$Locality) %>%
  group_by(Locality) %>%
  summarise(
    depth     = mean(`DepthInMeters (m)`, na.rm = TRUE),
    Island    = word(Locality, -1),
    latitude  = mean(`latitude (degrees_north)`, na.rm = TRUE),
    longitude = mean(`longitude (degrees_east)`, na.rm = TRUE)
  )

site_scores <- left_join(site_scores, site_meta, by = "Locality")

species_scores <- scores(nmds, display = "species") %>%
  as.data.frame() %>%
  rownames_to_column("ScientificName")

hulls <- site_scores %>%
  group_by(Island) %>%
  slice(chull(NMDS1, NMDS2))

ggplot() +
  geom_polygon(data = hulls,
               aes(x = NMDS1, y = NMDS2, fill = Island, group = Island),
               alpha = 0.3, color = NA) +
  geom_point(data = site_scores,
             aes(x = NMDS1, y = NMDS2, color = depth), size = 4) +
  geom_text(data = site_scores,
            aes(x = NMDS1, y = NMDS2, label = Locality),
            size = 3, vjust = 1) +
  scale_color_viridis(option = "plasma", name = "Mean Depth (m)") +
  geom_segment(data = species_scores,
               aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "darkgreen", linewidth = 0.5) +
  geom_text(data = species_scores,
            aes(x = NMDS1, y = NMDS2, label = ScientificName),
            color = "darkgreen", fontface = "italic", size = 3,
            hjust = 0.5, vjust = -0.4) +
  labs(
    title = str_wrap("Deep-Sea Glass Sponge Communities across the Hawaiian Archipelago", 48),
    subtitle = str_wrap("NMDS ordination | Bray-Curtis dissimilarity | sites colored by mean depth", 60),
    x = "NMDS1", y = "NMDS2"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

## Comparing NMDS and PCoA

When NMDS and PCoA arrange the sites in broadly similar configurations, it provides reassurance that
the pattern reflects genuine structure in the data rather than an artifact of any single method.

```{r nmds-pcoa-comparison}
nmds_sites <- scores(nmds, display = "sites") %>%
  as.data.frame() %>%
  rownames_to_column("Site")

pcoa_sites <- as.data.frame(pcoa$points[, 1:2]) %>%
  rownames_to_column("Site") %>%
  rename(PCo1 = V1, PCo2 = V2)

comparison_df <- left_join(nmds_sites, pcoa_sites, by = "Site")

plot_nmds <- ggplot(comparison_df, aes(x = NMDS1, y = NMDS2)) +
  geom_point(color = "blue", size = 3) +
  geom_text(aes(label = Site), vjust = -0.5, size = 3) +
  labs(title = "NMDS Ordination", x = "NMDS1", y = "NMDS2") +
  theme_minimal()

plot_pcoa <- ggplot(comparison_df, aes(x = PCo1, y = PCo2)) +
  geom_point(color = "darkgreen", size = 3) +
  geom_text(aes(label = Site), vjust = -0.5, size = 3) +
  labs(title = "PCoA Ordination", x = "PCo1", y = "PCo2") +
  theme_minimal()

plot_nmds + plot_pcoa
```

# Inferential Analysis: What Structures the Communities?

The preceding sections describe the data; this section tests specific hypotheses about it, using
methods designed to withstand scrutiny. The questions are addressed in approximate order of how
confidently the data can answer them:

1. Does depth structure both community composition and diversity? This is the primary question, and
   the one the dataset is best equipped to answer.
2. Which factors actually drive the differences between communities — depth, geography, or both — and
   could any apparent geographic pattern simply reflect the tendency of nearby sites to resemble one
   another?
3. After accounting for depth and the sampling-gear confound, does a genuine difference between
   islands remain, and does it persist under several robustness checks?
4. Which species are most characteristic of each island?

In short: depth is the well-supported result, whereas the island difference is weaker and
confounded. The analysis therefore leads with depth and treats any island effect as credible only if
it survives explicit sensitivity testing.

Two methodological decisions underpin everything that follows:

- **The dive is the unit of analysis, not the individual sponge.** Treating all 600-plus individual
  observations as independent would constitute pseudoreplication, since sponges recorded on the same
  dive are not independent of one another. Aggregating to roughly 33 dives provides valid
  replication.
- **Sampling effort and equipment are confounded.** Dives range from a single observation to well
  over a hundred, and different expeditions used different gear. Rather than ignore this, the
  analysis quantifies it explicitly and corrects for it.

## Sampling effort and confounds

```{r effort-summary}
station_data <- cleaned_sponges %>%
  filter(!is.na(Station), Station != "") %>%
  rename(Depth = `DepthInMeters (m)`,
         lat   = `latitude (degrees_north)`,
         lon   = `longitude (degrees_east)`)

# One row per dive: dominant island/gear, effort (n_obs), mean depth and coordinates.
station_meta <- station_data %>%
  group_by(Station) %>%
  summarise(
    Island     = names(which.max(table(Island))),
    Equipment  = names(which.max(table(SamplingEquipment))),
    n_obs      = n(),
    mean_depth = mean(Depth, na.rm = TRUE),
    latitude   = mean(lat,  na.rm = TRUE),
    longitude  = mean(lon,  na.rm = TRUE),
    .groups    = "drop"
  )

# Effort varies by orders of magnitude across dives:
summary(station_meta$n_obs)

# Island is partly confounded with sampling gear -- keep this in mind when
# interpreting any "island effect".
table(Island = station_meta$Island, Gear = station_meta$Equipment)
```

## The dive-level community matrix

```{r station-matrix}
station_matrix <- station_data %>%
  count(Station, ScientificName) %>%
  pivot_wider(names_from = ScientificName, values_from = n, values_fill = 0) %>%
  column_to_rownames("Station")

# Drop empty dives and align metadata to the matrix row order (critical for all
# downstream tests -- vegan returns results in matrix row order).
station_matrix <- station_matrix[rowSums(station_matrix) > 0, , drop = FALSE]
station_meta   <- station_meta %>%
  filter(Station %in% rownames(station_matrix))
station_meta   <- station_meta[match(rownames(station_matrix), station_meta$Station), ]

# Restrict island comparisons to islands with enough dives to support a test.
island_n <- table(station_meta$Island)
island_n
keep_islands <- names(island_n[island_n >= 3])

dim(station_matrix)
```

> A note on scope. Once observations are restricted to those identified to species and linked to a
> specific dive, nearly all of the usable data come from just two islands — Oahu and Hawaii.
> Accordingly, despite the word "archipelago" in the title, the inferential comparison is effectively
> Oahu versus Hawaii, with depth as a covariate. Stating this explicitly is preferable to
> overclaiming an archipelago-wide result.

## Diversity by dive

```{r diversity}
station_meta <- station_meta %>%
  mutate(
    richness = specnumber(station_matrix),
    shannon  = diversity(station_matrix, index = "shannon"),
    simpson  = diversity(station_matrix, index = "simpson")
  )

# Raw richness is effort-biased. Compute rarefied richness only for adequately
# sampled dives (>= min_effort observations) and rarefy to the smallest such dive
# so the comparison is fair. Tiny dives (e.g. a single observation) cannot yield a
# meaningful rarefied estimate and are left as NA.
min_effort <- 5
adequate   <- rowSums(station_matrix) >= min_effort   # aligns with station_meta rows
rare_size  <- min(rowSums(station_matrix)[adequate])

station_meta$rarefied_richness <- NA_real_
station_meta$rarefied_richness[adequate] <-
  as.numeric(rarefy(station_matrix[adequate, , drop = FALSE], sample = rare_size))

station_meta %>%
  select(Station, Island, n_obs, mean_depth, richness, rarefied_richness, shannon, simpson) %>%
  arrange(desc(shannon))
```

A guide to the columns: each row corresponds to one dive. *Richness* is the number of taxa recorded,
but it cannot be compared directly between dives, because a dive that recorded more sponges will
almost always accumulate more taxa. *Rarefied richness* corrects for this by estimating the number of
taxa expected if every dive had recorded the same number of individuals, and is therefore the
appropriate column for comparison. The *Shannon* and *Simpson* indices are standard diversity
measures that additionally reward an even distribution of individuals across taxa rather than
dominance by one. Rarefied richness is the measure carried forward into the depth–diversity analysis
below.

## The depth–diversity relationship

A classic question in deep-sea ecology is whether diversity peaks at intermediate depths or instead
increases or decreases steadily with depth. Rather than assume a particular form, the model below
compares a linear and a quadratic (unimodal) fit and lets the data indicate which is better
supported.

```{r depth-diversity, fig.cap="Effort-corrected (rarefied) species richness against mean dive depth, with the AIC-preferred trend line. A hump-shaped fit would indicate a mid-depth diversity peak; a flat or sloped line indicates a simpler or weak relationship. Point size shows sampling effort."}
# Raw richness is confounded with effort, so model rarefied richness vs depth
# (using only the adequately-sampled dives with a non-missing rarefied estimate).
div_df <- station_meta %>% filter(!is.na(rarefied_richness))

# Deep-sea diversity is classically unimodal (peaks at mid-bathyal depths), so we
# compare a linear and a quadratic fit rather than assuming a straight line.
mod_linear <- lm(rarefied_richness ~ mean_depth, data = div_df)
mod_quad   <- lm(rarefied_richness ~ poly(mean_depth, 2), data = div_df)

AIC(mod_linear, mod_quad)        # lower AIC = better-supported model
anova(mod_linear, mod_quad)      # is the quadratic term a significant improvement?

# Report whichever the data prefer.
best_form <- if (AIC(mod_quad) < AIC(mod_linear)) y ~ poly(x, 2) else y ~ x
summary(if (AIC(mod_quad) < AIC(mod_linear)) mod_quad else mod_linear)

ggplot(div_df, aes(x = mean_depth, y = rarefied_richness)) +
  geom_point(aes(color = Island, size = n_obs), alpha = 0.8) +
  geom_smooth(method = "lm", formula = best_form, se = TRUE,
              color = "grey30", linewidth = 0.8) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Depth–Diversity Relationship (effort-corrected)",
    subtitle = str_wrap(paste0("Rarefied richness at n = ", rare_size,
                      " obs/dive | fit = AIC-preferred model | point size = effort"), 60),
    x = "Mean dive depth (m)", y = "Rarefied species richness"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

## Rarefaction: accounting for unequal effort

Each curve represents one dive. A curve that flattens indicates that the dive recorded nearly all the
taxa it was likely to encounter, so its richness estimate is reliable. A curve still rising at its
endpoint indicates that the dive stopped before reaching saturation, so its richness represents a
lower bound rather than a complete count. These curves are the visual justification for relying on
the effort-corrected (rarefied) richness above.

```{r rarefaction, fig.height=5}
# Curves that have not plateaued indicate under-sampled dives -- a direct visual
# of why raw richness comparisons are unsafe here.
rarecurve(as.matrix(round(station_matrix)), step = 5, cex = 0.5, col = "steelblue",
          xlab = "Observations sampled", ylab = "Species detected",
          main = "Rarefaction Curves by Dive")
```

## PERMANOVA: testing composition by depth and island

This is the central test of the report. Because island, depth, and gear are confounded, simply asking
whether the islands differ would attribute to island any variation that depth or equipment actually
explains. To avoid this, the analysis uses PERMANOVA — a standard permutation-based test of community
composition — and fits all three factors simultaneously, evaluating each one after accounting for the
others (`by = "margin"`). Depth is entered as a quadratic term, since deep-sea gradients are
typically nonlinear. Under this design, the island term reflects only the variation it explains over
and above depth and gear.

```{r permanova}
# Restrict to islands with adequate replication for the grouped test.
keep_rows     <- station_meta$Island %in% keep_islands
comm_test     <- station_matrix[keep_rows, , drop = FALSE]
comm_test     <- comm_test[, colSums(comm_test) > 0, drop = FALSE]
meta_test     <- station_meta[keep_rows, ]
bray_test     <- vegdist(comm_test, method = "bray")

# Marginal test: each term controlling for the others. This is the honest test of
# an "island effect" given the gear/depth confounds.
set.seed(42)
adonis_margin <- adonis2(bray_test ~ Equipment + poly(mean_depth, 2) + Island,
                         data = meta_test, by = "margin", permutations = 999)
adonis_margin

# Sequential test (terms added in order) for comparison: shows how much depth
# explains on its own before island is considered.
set.seed(42)
adonis_seq <- adonis2(bray_test ~ poly(mean_depth, 2) + Equipment + Island,
                      data = meta_test, by = "terms", permutations = 999)
adonis_seq
```

Interpreting these tables: each row corresponds to one factor. The `R2` column gives the proportion
of community variation attributable to that factor, and `Pr(>F)` is its permutation p-value (with
0.05 the conventional threshold for significance). In the marginal table, the key row is `Island`,
which is evaluated after depth and gear have been accounted for; if it is small and non-significant,
no island difference remains once the sampling design is taken into account. The sequential table
enters depth first, illustrating how much of the variation depth explains on its own before the other
factors are considered.

```{r permanova-dispersion}
# PERMANOVA can be fooled by unequal within-group dispersion, so test that too.
# A non-significant betadisper result means a significant PERMANOVA reflects a
# genuine location (centroid) difference rather than a spread difference.
bd <- betadisper(bray_test, meta_test$Island)
set.seed(42)
permutest(bd)
```

This dispersion check is important for interpretation. PERMANOVA can register a "difference" when
groups differ only in how internally variable they are, rather than in their average composition. A
non-significant betadisper result indicates that the groups have comparable dispersion, which allows
a significant PERMANOVA result to be interpreted as a genuine difference in community type rather than
merely in variability.

```{r anosim}
# ANOSIM as an independent confirmation of group separation.
set.seed(42)
anosim(bray_test, grouping = meta_test$Island)
```

ANOSIM provides an independent assessment of the same question using a different statistic. Its `R`
ranges from 0 (groups indistinguishable) to 1 (groups completely separated); values near 0 indicate
that island membership conveys little information about a dive's community, even when the associated
p-value is small. Agreement between ANOSIM and the PERMANOVA result strengthens confidence in the
conclusion.

## envfit: Which Variables Drive the Ordination?

```{r station-nmds-envfit, fig.height=6, fig.cap="Dive-level NMDS with fitted environmental vectors. Arrow direction shows the gradient of each variable across the ordination and arrow length its strength; ellipses summarize each island. A long depth arrow indicates depth is a dominant driver. Vector significance is in the printed envfit output."}
set.seed(42)
nmds_station <- metaMDS(comm_test, distance = "bray", k = 2, trymax = 200, trace = 0)

ef <- envfit(nmds_station ~ mean_depth + latitude + longitude,
             data = meta_test, permutations = 999)
ef

# Build a ggplot ordination with fitted environmental vectors.
st_scores <- as.data.frame(scores(nmds_station, display = "sites"))
st_scores$Island <- meta_test$Island

vec_df <- as.data.frame(scores(ef, display = "vectors")) * ordiArrowMul(ef)
vec_df$var <- rownames(vec_df)

ggplot(st_scores, aes(NMDS1, NMDS2)) +
  stat_ellipse(aes(color = Island), linewidth = 0.7) +
  geom_point(aes(color = Island), size = 3, alpha = 0.85) +
  geom_segment(data = vec_df,
               aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),
               arrow = arrow(length = unit(0.25, "cm")), color = "grey20") +
  geom_text(data = vec_df, aes(x = NMDS1, y = NMDS2, label = var),
            color = "grey20", fontface = "bold", vjust = -0.4) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Station-Level NMDS with Environmental Vectors",
       subtitle = str_wrap(paste0("Bray-Curtis | stress = ", round(nmds_station$stress, 3),
                         " | vectors = envfit (p-values in printout above)"), 60)) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

## Robustness and sensitivity checks

A result from a single model specification warrants limited confidence. A genuine island difference
should persist under reasonable changes to the analysis. The marginal PERMANOVA is therefore rerun
three ways: with all taxa collapsed to genus (to test whether the result depends on inconsistent
identification depth); with abundances reduced to presence/absence using the Jaccard index (to test
whether it depends on observation counts and effort); and with the single largest dive removed (to
test whether it is driven by one influential sample). The relevant comparison is whether the `Island`
term remains stable across all three.

```{r robustness}
run_margin <- function(d, meta) {
  set.seed(42)
  adonis2(d ~ Equipment + poly(mean_depth, 2) + Island,
          data = meta, by = "margin", permutations = 999)
}

# (1) Genus-collapsed community matrix, aligned to the tested dives.
genus_matrix <- station_data %>%
  mutate(Genus = word(ScientificName, 1)) %>%
  count(Station, Genus) %>%
  pivot_wider(names_from = Genus, values_from = n, values_fill = 0) %>%
  column_to_rownames("Station")
genus_matrix <- genus_matrix[rownames(comm_test), , drop = FALSE]
genus_matrix <- genus_matrix[, colSums(genus_matrix) > 0, drop = FALSE]

cat("\n--- (1) Genus-collapsed (Bray-Curtis) ---\n")
run_margin(vegdist(genus_matrix, "bray"), meta_test)

# (2) Presence/absence using Jaccard.
cat("\n--- (2) Presence/absence (Jaccard) ---\n")
run_margin(vegdist(comm_test, method = "jaccard", binary = TRUE), meta_test)

# (3) Drop the single largest dive (most observations) and refit.
mega <- rownames(comm_test)[which.max(rowSums(comm_test))]
keep_nomega <- rownames(comm_test) != mega
comm_nomega <- comm_test[keep_nomega, , drop = FALSE]
comm_nomega <- comm_nomega[, colSums(comm_nomega) > 0, drop = FALSE]
cat("\n--- (3) Excluding largest dive:", mega, "---\n")
run_margin(vegdist(comm_nomega, "bray"), meta_test[keep_nomega, ])
```

The criterion here is consistency. If the island effect retains roughly the same magnitude and
significance across all three reruns, it can be regarded as robust. If it appears in one
specification but disappears under genus-level grouping, presence/absence, or removal of the largest
dive, it is most likely an artifact — and the appropriate conclusion is that no dependable island
difference exists once depth and sampling design are accounted for.

## Distance-decay: Mantel tests

An apparent island effect could also reflect geographic distance alone: nearby dives tend to support
similar communities simply because they are close together, regardless of island. Mantel tests
distinguish these possibilities by correlating community dissimilarity with distance matrices. The
decisive test is the partial Mantel: after controlling for depth, does geographic distance still
correlate with community dissimilarity? If it does not, then most of the apparent geographic
structure is attributable to depth expressed across space.

```{r mantel}
# Great-circle (haversine) geographic distance in km between dives.
haversine_dist <- function(lon, lat) {
  R <- 6371; lon <- lon * pi/180; lat <- lat * pi/180
  n <- length(lon); m <- matrix(0, n, n)
  for (i in seq_len(n)) for (j in seq_len(n)) {
    dlon <- lon[j] - lon[i]; dlat <- lat[j] - lat[i]
    a <- sin(dlat/2)^2 + cos(lat[i]) * cos(lat[j]) * sin(dlon/2)^2
    m[i, j] <- 2 * R * asin(sqrt(a))
  }
  as.dist(m)
}

geo_dist   <- haversine_dist(meta_test$longitude, meta_test$latitude)
depth_dist <- dist(meta_test$mean_depth)

cat("\n--- Community vs DEPTH distance (is there a depth gradient?) ---\n")
set.seed(42); mantel(bray_test, depth_dist, permutations = 999)

cat("\n--- Community vs GEOGRAPHIC distance (distance-decay?) ---\n")
set.seed(42); mantel(bray_test, geo_dist, permutations = 999)

cat("\n--- Partial Mantel: geography controlling for depth ---\n")
set.seed(42); mantel.partial(bray_test, geo_dist, depth_dist, permutations = 999)
```

Interpreting the results: each test reports a correlation (`r`) and a p-value. A positive
community–depth correlation supports a depth gradient, while a positive community–distance
correlation would indicate distance-decay (nearby dives resembling one another). The decisive result
is the partial Mantel: if the geographic correlation falls toward zero once depth is controlled for,
the apparent geographic pattern is largely depth expressed across space — again pointing to depth as
the underlying driver.

## Indicator species by island

```{r indicator-species}
# Indicator Species Analysis (IndVal): which species are characteristic of each
# island? Requires the 'indicspecies' package; skipped gracefully if absent.
if (requireNamespace("indicspecies", quietly = TRUE)) {
  set.seed(42)
  indval <- indicspecies::multipatt(
    comm_test, meta_test$Island,
    func = "IndVal.g", control = permute::how(nperm = 999)
  )
  summary(indval, indvalcomp = TRUE)
} else {
  message("Install 'indicspecies' (and 'permute') to run indicator species analysis.")
}
```

Each taxon listed is statistically associated with a particular island or island group. The
indicator statistic (closer to 1 denotes a stronger, more exclusive association) and its p-value (low
values indicate the association is unlikely to be due to chance) convey how reliable each association
is. This analysis complements the PERMANOVA: even where the overall island difference is weak,
individual taxa may still be characteristic of one island. Such associations are informative
natural-history detail, but should be interpreted in light of the small number of dives involved.

## Community heatmap

```{r heatmap, fig.height=7, fig.cap="Community matrix as a heatmap: each cell is the (log) count of a taxon on a dive, with dives grouped by island. Vertical bands of color reveal taxa that dominate across many dives; gaps reveal taxa restricted to particular dives or islands."}
heat_df <- station_matrix %>%
  as.data.frame() %>%
  rownames_to_column("Station") %>%
  pivot_longer(-Station, names_to = "Species", values_to = "count") %>%
  left_join(station_meta %>% select(Station, Island), by = "Station")

ggplot(heat_df, aes(x = Species, y = Station, fill = log1p(count))) +
  geom_tile(color = "grey92") +
  facet_grid(Island ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis(option = "mako", name = "log(count + 1)") +
  labs(title = "Glass Sponge Community Matrix (dive × species)",
       x = NULL, y = "Dive (Station)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
        plot.title = element_text(face = "bold"))
```

## Spatial distribution of dives

The dives are plotted on the Hawaiian seafloor itself. When the `marmap` package is available, the
basemap is real NOAA bathymetry (with depth contours), which puts each dive in its actual physical
setting — on the island slopes and offshore terraces where glass sponges occur. Points are colored by
island and sized by sampling effort, making the two-island concentration of well-sampled dives
(Oahu and Hawaii) immediately apparent. If `marmap` is not installed, the code falls back to a simple
coastline outline so the document still knits.

```{r map, fig.height=6.5, fig.cap="Dive locations across the Main Hawaiian Islands, shown on NOAA bathymetry where available. Each point is one dive, colored by island and sized by sampling effort."}
map_pts <- station_data %>%
  group_by(Station) %>%
  summarise(lon = mean(lon, na.rm = TRUE), lat = mean(lat, na.rm = TRUE),
            depth = mean(Depth, na.rm = TRUE), n = n(),
            Island = names(which.max(table(Island))), .groups = "drop")

xlim <- c(-160.5, -154.5); ylim <- c(18.5, 22.5)

# Try to download/cache NOAA bathymetry for a true seafloor basemap.
bathy <- NULL
if (requireNamespace("marmap", quietly = TRUE)) {
  bathy <- tryCatch(
    marmap::getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                          lat1 = ylim[1], lat2 = ylim[2],
                          resolution = 2, keep = TRUE),
    error = function(e) NULL)
}

if (!is.null(bathy)) {
  bathy_df <- marmap::fortify.bathy(bathy)

  ggplot() +
    geom_raster(data = bathy_df, aes(x = x, y = y, fill = z), show.legend = FALSE) +
    geom_contour(data = bathy_df, aes(x = x, y = y, z = z),
                 breaks = c(-500, -1000, -2000, -3000, -4000),
                 colour = "grey75", linewidth = 0.2) +
    marmap::scale_fill_etopo(guide = "none") +
    geom_point(data = map_pts, aes(x = lon, y = lat, colour = Island, size = n), alpha = 0.9) +
    scale_colour_brewer(palette = "Set1", name = "Island") +
    scale_size_continuous(name = "Observations\n(effort)") +
    coord_quickmap(xlim = xlim, ylim = ylim) +
    labs(title = "Glass Sponge Dive Locations on the Hawaiian Seafloor",
         subtitle = str_wrap("Basemap: NOAA bathymetry with depth contours | points = dives, colored by island and sized by effort", 70),
         x = "Longitude", y = "Latitude") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))
} else {
  # Fallback: simple coastline outline (no bathymetry).
  p_map <- ggplot(map_pts, aes(lon, lat))
  if (requireNamespace("maps", quietly = TRUE)) {
    p_map <- p_map + borders("world", xlim = xlim, ylim = ylim,
                             fill = "grey88", colour = "grey70")
  }
  p_map +
    geom_point(aes(colour = Island, size = n), alpha = 0.85) +
    scale_colour_brewer(palette = "Set1", name = "Island") +
    scale_size_continuous(name = "Observations") +
    coord_quickmap(xlim = xlim, ylim = ylim) +
    labs(title = "Glass Sponge Dive Locations, Main Hawaiian Islands",
         subtitle = str_wrap("Install 'marmap' for a bathymetric basemap | points sized by sampling effort", 70),
         x = "Longitude", y = "Latitude") +
    theme_minimal(base_size = 11) +
    theme(plot.title = element_text(face = "bold"))
}
```

# Key Findings

Observations spanning the full cleaned dataset:

- Glass sponges were recorded from approximately 90 m to over 2,700 m, with the majority of records
  concentrated in the upper 1,000 m.
- A small number of taxa account for most observations (notably *Regadrella*, *Ijimalophus
  hawaiicus*, and *Farrea occa*), and these taxa show clear depth partitioning.

The primary result — depth:

- Depth is the clearest and best-supported driver of community structure. It is confirmed by three
  complementary analyses that point in the same direction: a gradient term in the PERMANOVA, a fitted
  vector in the ordination, and a Mantel correlation of community dissimilarity with depth. For
  diversity, an AIC comparison determines whether a linear or unimodal relationship is better
  supported.
- Diversity is reported on an effort-corrected (rarefied) basis throughout, because dives differed
  greatly in effort and raw counts would favor the most intensively sampled dives.

The secondary result — islands, interpreted conditionally:

- The comparison is at most an Oahu-versus-Hawaii contrast, as these are the only two islands with
  sufficient records to test; characterizing the result as archipelago-wide would overstate the data.
- Island identity is confounded with sampling gear and depth, so the only defensible test evaluates
  what island explains after those factors are accounted for.
- Even then, a genuine island difference must persist under the robustness checks (genus-level
  grouping, presence/absence, removal of the largest dive) and be distinguishable from distance-decay.
  If it does not, the appropriate conclusion is that no island effect remains once depth and sampling
  design are taken into account.
- With roughly 12 dives on one island and 21 on the other, statistical power is low; a non-significant
  result indicates insufficient evidence rather than a demonstrated absence of difference.

In summary, this report presents its conclusions together with their associated uncertainty rather
than as unqualified claims. The well-supported finding is that depth structures these communities;
the island result is reported conditionally and only to the extent the data permit.

# Discussion

Taken together, the analyses are mutually consistent. The clearest and most reproducible signal in
the data is depth: the sponges occupy distinct depth bands, depth registers as a community gradient
in both the ordination and the distance-based tests, and it carries substantial weight in the
PERMANOVA before any other factor is considered. It is also the conclusion the data are best
positioned to support, since the observations span the depth range densely and depth is measured
directly for every record.

The island question is more tenuous, and the analysis is designed to treat it accordingly. Because
island, gear, depth, and geographic distance are interrelated, an uncritical comparison would
attribute genuine differences to island when another factor is responsible. Testing island only after
gear and depth, repeating the test under three alternative specifications, and asking whether distance
alone could account for the pattern collectively place the burden of proof where it belongs.
Whichever way the results fall, the conclusion is defensible: if the island effect survives all of
these checks, it is a credible secondary finding; if it does not, then no island difference exists
beyond depth and the structure of the sampling.

This two-tier outcome — a confident conclusion regarding depth and a carefully bounded one regarding
geography — is itself the central result. It reflects what opportunistic deep-sea exploration data can
and cannot support, and it is more useful than an overstated claim of archipelago-wide biogeographic
structure.

# Limitations

- **Observational rather than standardized data.** The records derive from opportunistic ROV and
  submersible dives, so detection depends on dive duration, visibility, and substrate. Abundances are
  observation counts rather than density estimates, which is why composition is cross-checked with
  presence/absence (Jaccard) and richness is rarefied rather than compared directly.
- **Confounded sampling design.** Island, sampling gear, and expedition are correlated and only
  partially separable, so even the marginal island term has limited interpretability and should be
  read as variation beyond depth and gear rather than a clean biogeographic effect.
- **Low statistical power.** The community comparison rests on roughly 12 dives (Hawaii) and 21 dives
  (Oahu) and roughly two dozen taxa, several of them rare. A non-significant result indicates insufficient
  evidence rather than a demonstrated absence, and effect-size estimates are correspondingly
  uncertain.
- **Mixed taxonomic resolution.** Identifications range from genus to subspecies. The genus-level
  robustness analysis tests whether conclusions depend on this, but some finer-scale ecological
  structure may be obscured regardless.
- **Two-island scope.** After quality filtering, only Oahu and Hawaii retain sufficient
  dive-referenced records, so the results should not be generalized to the wider archipelago.

```{r session-info}
# Recorded for reproducibility.
sessionInfo()
```
